---
title: "ACS NPS Simulation Summary"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE)
library(dplyr)
library(knitr)
```

## Model-level summary (ordered by MSE)

```{r}
load("data/ACS_NPS_simulation_results.RData")
stopifnot(exists("summary_df"))

# Wall-clock duration of full 15-sim run
wall_time <- list(
  start = as.POSIXct("2026-01-20 20:11:00", tz = Sys.timezone()),
  end   = as.POSIXct("2026-01-21 08:08:06", tz = Sys.timezone())
)
wall_elapsed <- difftime(wall_time$end, wall_time$start, units = "hours")
cat(sprintf("Full run wall time: %.2f hours (start %s, end %s)\n\n",
            as.numeric(wall_elapsed),
            format(wall_time$start, \"%Y-%m-%d %H:%M\"),
            format(wall_time$end, \"%Y-%m-%d %H:%M\")))

summary_df <- summary_df %>%
  arrange(MSE)

kable(
  summary_df,
  digits = c(5, 5, 5, 3, 3),
  caption = "Model metrics aggregated across simulations"
)
```

## Highlights
- **Top MSE:** bulm_uncond, bulm_beta, mrp-p, mrp-p-int (all ≲0.0011).
- **Coverage leaders:** bulm_ps_only (0.957), NPS prior PP (0.955–0.950), mrp-r variants (~0.93).
- **Interval Score (lower is better):** bulm_uncond/bulm_beta ≈0.15–0.16 are best; IPW HT (beta) is worst (3.006).
- **Lowest performers (MSE):** IPW HT (beta) ≫ direst > VSW.

## Notes
- VSW reports NA for interval metrics by design.
- NPS power-prior variants now produce intervals (pseudolikelihood and raking weight-covariate).
